---
title: "FPL_PCA"
author: "Nima Hamedani-Raja"
date: "28/11/2020"
output: 
  html_document: 
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# about

Here, I am (back)documenting (some of) the steps I've taken to use *PCA* to analyse *FPL* teams.  
To do: explain template and differentials, PCA, ... 

```{r init}
source("Functions/init.R")
```

# Getting data

## Data

I am using *fplscrapR*. I have also modified some of fplscrapR's methods.  
I am focusing on two subsets of data: teams of 8 players who compete in a private league, as well as the top 10K players.  
I am focusing on week 9 rather arbitrarily, simply because that's when I did this!

## getting and cleaning teams

### getting all players of all teams of a league

I'm using *get_a_league_teams()*.

```{r get_a_league_teams}
writeLines(readLines("Functions/get_a_league_teams.R"))

private_league_entries_teams_flat <- get_a_league_teams(1138028,9)
head(private_league_entries_teams_flat) %>% t() %>% kable()
```

### Cleaning private_league_entries_teams_flat

Not all these columns are needed. Let's trim it down.

```{r trim_private_league_entries_teams_flat}
private_league_entries_teams_flat <- private_league_entries_teams_flat %>% 
    select(id, playername,multiplier,entry,player_name,entry_name,league_name,week_number,
           league_name,total_points,value)
tail(private_league_entries_teams_flat) %>% kable()
```

Multiplier (used by *FPL API*) demonstrates wether a player was benched (0), played (1), or was captain (2). I'm interested in active teams. so I'll remove the unused subs.

```{r remove_multiplier_0}
private_league_entries_teams_flat <- private_league_entries_teams_flat %>% filter(multiplier != 0)
```

### dcating private_league_entries_teams_flat

After dcasting, each team is reprented by a vector of 1s(10 player), 2 (the captain), and 0s (the rest).  
```{r dcast_private_league_entries_teams_flat}
private_league_entries_teams <- 
    private_league_entries_teams_flat %>% dcast(entry_name ~ playername, value.var = "multiplier")

private_league_entries_teams[is.na(private_league_entries_teams)] <- 0

row.names(private_league_entries_teams) <- private_league_entries_teams$entry_name

private_league_entries_teams <- private_league_entries_teams %>% select(-entry_name)

t(private_league_entries_teams)
```

# PCA  

## PCA  

Let's PCA!  

```{r private_league_PCA}
private_league_PCA <- prcomp(private_league_entries_teams)

private_league_PCA$sdev
```

unfortunately, all sdev-s are significant. A significant chunk of information will be lost if some of them are cut out.  
Let's make sure we can reconstruct the original matrix:  

```{r private_league_PCA test}
(((private_league_PCA$x %*% t(private_league_PCA$rotation)) + 
    (as.matrix(rep(1,8)) %*% t(private_league_PCA$center))) - 
   private_league_entries_teams) %>% 
  melt() %>% 
  summarise(min(value), max(value))
```

## Principal Axes  

Which players emerge as important in Principal Axes?

```{r private_league_PCA - which players in PA}
private_league_PCA$rotation %>% 
  melt() %>% 
  filter(abs(value) > 0.3) %>% 
  dcast(Var1 ~ Var2, value.var = "value") %>%
  kable()
```

Owning Fernandes and Kane  seem to contribute a lot into the differences between teams (but not as much I'd like). 

To reconstruct the teams using PAs, the following rotation matrix can be used.
```{r private_league_PCA - rotation matrix}
private_league_PCA$x %>% kable()
```

## Player's points contributions

How many points does a player contribute to a PA?  

```{r private_league_players_PCA_points}
private_league_players_PCA_points_flat <- 
  inner_join( 
    private_league_PCA$rotation %>% 
      melt() %>% 
      mutate(Var1 = as.character(Var1)) %>%
      rename(playername = Var1, PA = Var2),
    private_league_entries_teams_flat %>% 
      select(playername, total_points) %>% 
      unique(),
    by = "playername"
  ) %>%
  group_by(playername, PA) %>%
  summarise(points = sum(value * total_points)) %>%
  ungroup() 

private_league_players_PCA_points <- 
  private_league_players_PCA_points_flat %>%
  dcast(playername ~ PA, value.var = "points") 

private_league_players_PCA_points %>% kable()
```

Unfortunately, they are NOT insignificant. 

As a test, let's find game week points using this matrix and compare it with actual. 

```{r entries_PA_points - test}
entries_PA_points <- 
  data.frame(points = 
  private_league_PCA$x %*% 
  (private_league_players_PCA_points %>% select(-playername) %>% t()) %>% 
  rowSums())

entries_PA_points <-
  entries_PA_points %>% 
  mutate(entry_name = row.names(entries_PA_points))
average_team_points <- 
  inner_join(
    data.frame(value = private_league_PCA$center) %>% 
      mutate(playername = row.names(data.frame(private_league_PCA$center))), 
    private_league_entries_teams_flat %>% 
      select(playername, total_points) %>% 
      unique(),
    by = "playername") %>% 
  summarise(average_team_points = sum(value*total_points)) %>% 
  as.numeric()

entries_points_1 <- entries_PA_points %>% mutate(points = points + average_team_points)

entries_points_2 <-
  private_league_entries_teams_flat %>% 
  group_by(entry_name) %>%
  summarise(points = sum(total_points*multiplier))

inner_join(entries_points_1,entries_points_2,by = "entry_name") %>% kable()
```

## Summary

It didn't work (expand later)
